A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
more

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{In practice, it can be useful to specify the same selector
for several constructors of the same (co)datatype,
as long as the argument types coincide \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is not allowed by SMT-LIB 2.}
Schematically, using a syntax similar to Standard ML, Haskell,
or SMT-LIB:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{111}{:}\bigr]\; \tau_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \tau_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \tau_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \tau_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \tau_{ij1}\times\cdots\times\tau_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \tau_{ijk}$. Defaults are assumed for
the selector names if they are omitted.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{pllist}(\alpha) & \,=\, \const{PLNil} \,\mid\, \const{PLCons}(\alpha,\, \ty{pllist}(\alpha)) \\[-.5\jot]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{pllist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here. In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case. So the only
genuinely interesting cases missing are mixed nested (co)recursion
(as in the \ty{tree} example above) as well as (co)recursion through a
non-(co)datatype (both of which make sense \cite{blanchette-et-al-2014-codata}).

@@

Datatypes can be defined semantically as the initial model of an equational
specification of the selector--constructor equations. This is the view followed
by Barrett et al.\ \cite{barrett-et-al-2010}. A disadvantage with this approach
is that it does not naturally account for selectors applied to wrong
constructors. Barrett et al.\ address this by parameterizing the construction by
default values, but these give rise to spurious equalities between unrelated
terms (e.g., $hd(nil1) \teq hd(nil2)$). This could be corrected, but the added
compexity seems to suggest that selectors are better specified axiomatically.

  * final coalgebra
    * this time, no condition: always exists
    e.g. the specification

      codatatype x = X x

    gives rise to a singleton type {X (X (X ...))}.

  * discriminators/selectors for initial algebras
  * constructors for final coalgebras

  * Introduce the discriminators already here.

  * ODDITY: Barrett et al. overspecify things (not good w.r.t. SMT-LIB 2, or
    other solvers, e.g., Z3)

@@

>>>>

\newcommand\vthinspace{\kern+0.083333em}
\newcommand\negvthinspace{\kern-0.083333em}

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{In practice, it can be useful to specify the same selector
for several constructors of the same (co)datatype,
as long as the argument types coincide \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is not allowed by SMT-LIB 2.}
Schematically, using a syntax similar to Standard ML, Haskell,
or SMT-LIB:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{gllist}(\alpha) & \,=\, \const{GLNil} \,\mid\, \const{GLCons}(\alpha,\, \ty{gllist}(\alpha)) \\[-.5\jot]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{gllist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here. In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case. So the only
genuinely interesting cases missing are mixed nested (co)recursion
(as in the \ty{tree} example above) as well as (co)recursion through a
non-(co)datatype (both of which make sense \cite{blanchette-et-al-2014-codata}).

@@

@@

Datatypes can be defined semantically as the initial model of an equational
specification of the selector--constructor equations. This is the view followed
by Barrett et al.\ \cite{barrett-et-al-2010}. A disadvantage with this approach
is that it does not naturally account for selectors applied to wrong
constructors. Barrett et al.\ address this by parameterizing the construction by
default values, but these give rise to spurious equalities between unrelated
terms---for example, $\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$.
This could be corrected, but the added
compexity seems to suggest that selectors are better specified axiomatically.
